---
title: "Target Trial Simulation"
author: "Susan Hoffman, MPH"
format: 
  pdf:
    include-in-header:
      text: |
        \usepackage{multirow}
editor: visual
---

# Simulation Parameters

Exposure:

-   The primary causal question of interest focuses on antibiotic initiation at or after 24 weeks gestation

Outcome

-   Preterm delivery

Causal contrast:

-   Intention to treat effect

Exchangeability considerations:

-   Age (\<18 - 4%; 18-35 - 79%; \>=35 - 17%)

-   Parity (0 - 41%, \>=1 - 59%)

-   HIV status (Negative - 78%; Living with HIV - 22%)

# Data generation

This synthetic data set was simulated based on Caniglia, et al. (2023) and is not representative of a real data set or real causal effects. This data is meant to be used for this sandbox example only and will not match what was seen in the manuscript.

```{r dataSim, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
set.seed(325)

# Sample Size
n <- 111403

# Baseline Covariates 

## Age from a Normal dist. with a min of 18 and a max of 45 
age <- sample(c("<18", "18-35", ">35"), 
              n, 
              replace = T, 
              prob = c(4, 79, 17))

## Parity from a binomial dist. where parity = 1 has a prob of 0.6
parity <- rbinom(n, size = 1, prob = 0.6)

## HIV from a binomial dist. where hiv = 1 has a prob of 0.22
hiv <- rbinom(n, size = 1, prob = 0.22)

covs <- cbind(age = age, parity = parity, hiv = hiv)
covs <- as.data.frame(covs)
covs$participant_id <- rownames(covs)

# Simulate the exposure 

## Exposure probabilities by week 
probs <- c(1371/110032,
           1382/108459,
           1310/109917,
           1400/105239,
           1326/103623,
           1403/101839,
           1363/100090,
           1376/98143,
           1256/96226,
           1231/94100,
           1227/91797,
           1208/88846,
           1156/85324)

# Generating exposure data matrix 
exposure_data <- matrix(0, nrow = n, ncol = length(probs))

for (i in seq_along(probs)) {
  # Determine individuals who were previously unexposed
  unexposed_individuals <- which(exposure_data[, i] == 0)
  
  # Generate exposure data for individuals who were previously unexposed
  if (length(unexposed_individuals) > 0) {
    exposure_data[unexposed_individuals, i] <- sample(c(0, 1), 
                                                      length(unexposed_individuals),
                                                      replace = TRUE, 
                                                      prob = c(1 - probs[i], probs[i]))
  }
  
  # Setting exposure to NA for individuals who were previously exposed
  if (i > 1) {
    previously_exposed <- exposure_data[, i - 1] == 1
    exposure_data[previously_exposed, i:length(probs)] <- NA
  }
}

# simulate the outcome 

## Outcome probabilities by week 
probs_o_0 <- c(
  15672/110032,
  15233/108549,
  14802/106917,
  14306/105239,
  13808/103623,
  13208/101839,
  12617/100090,
  11869/98143,
  11031/96226,
  9990/94100,
  8769/91797,
  6923/88846,
  4551/85324
)

probs_o_1 <- c(
  219/1371,
  248/1382,
  199/1310,
  218/1400,
  208/1326,
  219/1403,
  205/1363,
  177/1376,
  177/1256,
  146/1231,
  145/1227,
  103/1208,
  6/1156
)

# Simulate the outcome 
outcome_data <- matrix(0, nrow = n, ncol = length(probs_o_0))

for (i in seq_along(probs_o_0)) {
  # Determine individuals without outcome
  atrisk_individuals <- which(outcome_data[, i] == 0)
  
  # Generate outcome data among those at risk 
  if (length(atrisk_individuals) > 0) {
    # Apply different probabilities based on exposure status
    outcome_probs <- ifelse(
      exposure_data[atrisk_individuals, i] == 1, probs_o_1[i], probs_o_0[i]
      )
    
    outcome_probs <- outcome_probs[!is.na(outcome_probs)]
    
    # Generate outcome data for unexposed individuals
    outcome_data[atrisk_individuals, i] <- sample(c(0, 1), 
                                                      length(atrisk_individuals),
                                                      replace = TRUE, 
                                                      prob = c(1 - outcome_probs[i], outcome_probs[i]))
  }
  
  # Setting outcome to NA for individuals who have the outcome
  if (i > 1) {
    previously_exposed <- outcome_data[, i - 1] == 1
    outcome_data[previously_exposed, i:length(probs_o_0)] <- NA
  }
}

# Adding column names 
colnames(outcome_data) <- paste0("wk_", 24:36)

# Adding column names 
exposure_data <- as.data.frame(exposure_data)
outcome_data <- as.data.frame(outcome_data)

start <- 24
end <- 36
col_names <- paste0("wk_", as.character(start:end))

colnames(exposure_data) <- col_names
colnames(outcome_data) <- col_names

# Adding participant ID 
exposure_data$participant_id <- rownames(exposure_data)
outcome_data$participant_id <- rownames((outcome_data))


# Combining into final dataset 
exposure_long <- pivot_longer(exposure_data, 
                              cols = starts_with("wk"), 
                              names_to = "week", 
                              values_to = "exposure_info")

outcome_long <- pivot_longer(outcome_data, 
                              cols = starts_with("wk"), 
                              names_to = "week", 
                              values_to = "outcome_info")

final_data <- exposure_long
final_data <- cbind(final_data, outcome = outcome_long$outcome_info)
ids <- final_data$participant_id

final_data <- merge(covs, final_data, by = "participant_id")
final_data$participant_id <- ids


# count <- final_data %>% 
#   filter(week == "wk_35" & exposure_info == 1) %>% 
#   nrow()
```

# Analysis

```{r analysis, message=FALSE, warning=FALSE}
# Getting number of exposed participants per week 
exposure_counts <- final_data %>%
  group_by(week, exposure_info) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = exposure_info, 
              values_from = count, 
              names_prefix = "exposure_") %>%
  mutate(exposure_1 = coalesce(exposure_1, 0), 
         exposure_0 = coalesce(exposure_0, 0))

# Getting the number of outcome events per week by exposure status 
outcome_counts <- final_data %>%
  group_by(week, exposure_info) %>%
  summarise(outcome_events = sum(outcome, na.rm = TRUE)) 

# Data cleaning - creating factor variables

factor_names <- c('age', 'hiv', 'parity')

final_data[, factor_names] <- lapply(final_data[, factor_names], factor)

head(final_data)

# Data cleaning - creating numeric week variable 

final_data$week_num <- as.numeric(gsub("wk_", "", final_data$week))

# adjusted models by week 

# Create a list to store results for each week
adj_model_results <- list()

for (wk in paste0("wk_", 24:36)) {
  # Subset data for the current week
  data_wk <- final_data %>% filter(week == wk)
  
  # Fit log-binomial regression model and store results
  adj_model_results[[wk]] <- glm(outcome ~ exposure_info +
                                   hiv +
                                   parity +
                                   age, 
                                 data = data_wk, 
                                 family = binomial(link = "log")) # log link for RR
}


# Pull RR (exp. beta values) and 95% confidence intervals into a dataframe 
## Create a null dataframe 
results_adj <- data.frame(
  Week = character(),
  Coefficient = numeric(),
  Lower_CI = numeric(),
  Upper_CI = numeric(),
  stringsAsFactors = FALSE
)

for (wk in paste0("wk_", 24:36)) {
  # Extract coefficients and confidence intervals
  coefficients <- exp(adj_model_results[[wk]]$coefficients)
  ci <- exp(confint(adj_model_results[[wk]]))
  
  # Store results in the data frame
  results_adj <- rbind(results_adj, data.frame(
    Week = wk,
    Coefficient = coefficients["exposure_info"],
    Lower_CI = ci["exposure_info", "2.5 %"],
    Upper_CI = ci["exposure_info", "97.5 %"]
  ))
}

# Pooled results 

library(splines)

model_pooled <- glm(outcome ~ exposure_info +
                      hiv +
                      parity +
                      age +
                      ns(week_num, df = 2), # adding flexibility to the model for GA 
                    data = final_data, 
                    family = binomial(link = "log")) # log link for RR

co_pooled <- exp(model_pooled$coefficients)

# need to bootstrap confidence intervals 
replicates <- 5 # this needs to be increased for a real analysis 
bootstrap_estimates <- NULL

for (i in 1:replicates) {
  # set seed to i so we get a unique sample each time 
  set.seed(i)
  
  # sample the data
  index <- sample(1:nrow(final_data), nrow(final_data), replace = T)
  
  final_data_sample <- final_data[index,]
  
  # estimate the association in the resample 
  model_boot <- glm(outcome ~ exposure_info +
                      hiv +
                      parity +
                      age +
                      ns(week_num, df = 2), # adding flexibility to the model for GA 
                    data = final_data_sample, 
                    family = binomial(link = "log")) # log link for RR
  
  bootstrap_estimates <- rbind(
    bootstrap_estimates,
    coef(model_boot)[2]
  )
  
  # Print iteration number when i is divisible by 10
  if (i %% 10 == 0) {
    print(paste("Iteration:", i))
  }
  
}

se_estimate <- sd(bootstrap_estimates)

lcl_pooled <- exp(coef(model_pooled)[2] - 1.96 * se_estimate)
ucl_pooled <- exp(coef(model_pooled)[2] + 1.96 * se_estimate)

pooled_dat <- data.frame(
  Week = "pooled", 
  Coefficient = co_pooled[2],  
  Lower_CI = lcl_pooled, 
  Upper_CI = ucl_pooled
)

results_adj <- rbind(results_adj, pooled_dat)
```

# Tables and Figures

## RR summary figure

```{r fig, message=FALSE, warning=FALSE}
library(ggplot2)

ggplot(results_adj, aes(x = Week, 
                        y = Coefficient)) +
  geom_pointrange(aes(ymin = Lower_CI, 
                      ymax = Upper_CI), 
                  color = "black") +
  geom_point(color = "black", 
             size = 3) +
  geom_hline(yintercept = 1, 
             linetype = 'solid',
             color = 'red') +
  labs(x = 'Week', 
       y = "Relative Risk",
       title = "Risk of PTB") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))

```

## Table

To generate Latex tables online: <https://www.tablesgenerator.com/latex_tables#>

```{=tex}
\begin{table}[]
\begin{tabular}{lllll}
\hline
Antibiotic Initiation at:             & Initiation & \# of individuals & Events, \# & RR (95\% CI)                                                                  \\ \hline
\multirow{2}{*}{24 weeks}             & No         & 110082            & 15659      & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}0.99 \\ (0.86, 1.12)\end{tabular}} \\
                                      & Yes        & 1321              & 186        &                                                                               \\
\multirow{2}{*}{25 weeks}             & No         & 108628            & 13186      & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}1.06\\ (0.92, 1.21)\end{tabular}}  \\
                                      & Yes        & 1454              & 186        &                                                                               \\
\multirow{2}{*}{26 weeks}             & No         & 107316            & 10975      & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}0.99\\ (0.84, 1.16)\end{tabular}}  \\
                                      & Yes        & 1312              & 133        &                                                                               \\
\multirow{2}{*}{27 weeks}             & No         & 105876            & 8963       & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}0.99\\ (0.84, 1.17)\end{tabular}}  \\
                                      & Yes        & 1440              & 124        &                                                                               \\
\multirow{2}{*}{28 weeks}             & No         & 104555            & 7679       & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}0.98\\ (0.80, 1.17)\end{tabular}}  \\
                                      & Yes        & 1321              & 95         &                                                                               \\
\multirow{2}{*}{29 weeks}             & No         & 103095            & 6289       & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}0.96\\ (0.79, 1.17)\end{tabular}}  \\
                                      & Yes        & 1460              & 85         &                                                                               \\
\multirow{2}{*}{30 weeks}             & No         & 101656            & 5265       & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}1.28\\ (1.06, 1.52)\end{tabular}}  \\
                                      & Yes        & 1439              & 100        &                                                                               \\
\multirow{2}{*}{31 weeks}             & No         & 100249            & 4446       & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}1.24\\ (0.99, 1.51)\end{tabular}}  \\
                                      & Yes        & 1407              & 74         &                                                                               \\
\multirow{2}{*}{32 weeks}             & No         & 98962             & 3636       & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}1.06\\ (0.80, 1.35)\end{tabular}}  \\
                                      & Yes        & 1287              & 50         &                                                                               \\
\multirow{2}{*}{33 weeks}             & No         & 97722             & 2972       & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}0.91\\ (0.65, 1.24)\end{tabular}}  \\
                                      & Yes        & 1240              & 34         &                                                                               \\
\multirow{2}{*}{34 weeks}             & No         & 1317              & 2327       & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}0.74\\ (0.49, 1.06)\end{tabular}}  \\
                                      & Yes        & 96405             & 24         &                                                                               \\
\multirow{2}{*}{35 weeks}             & No         & 1303              & 1691       & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}1.07\\ (0.72, 1.52)\end{tabular}}  \\
                                      & Yes        & 95102             & 26         &                                                                               \\
\multirow{2}{*}{36 weeks}             & No         & 93841             & 1037       & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}0.62\\ (0.29, 1.15)\end{tabular}}  \\
                                      & Yes        & 1261              & 8          &                                                                               \\ \hline
\multirow{2}{*}{Pooled (24-36 weeks)} & No         & 1323489           & 84125      & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}1.02\\ (0.98, 1.07)\end{tabular}}  \\
                                      & Yes        & 17562             & 1125       &                                                                               \\ \hline
\end{tabular}
\end{table}
```
